---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

# ukbFGSEA

<!-- badges: start -->
<!-- badges: end -->

## Overview

The goal of ukbFGSEA is to apply gene set enrichment analysis, based on FGSEA package, to Genebass data released by Karczewski et al. The package is amed to help researchers to perform detailed evaluations of gene set enrichment across a wide range of phenotypes. 
The package mainly includes three functions:

- `ukbfgsea()` applies `fgsea::fgseaMultilevel()` - an adaptive multilevel splitting Monte Carlo approach, to a single phenotype in the UK Biobank Genebass dataset for a specified input gene set.

- `map_ukbfgsea()` applies `fgsea::fgseaMultilevel()` to multiple phenotypes in the UK Biobank Genebass dataset for a specified input gene set. This function is suitable for large-scale(or full) Genebass dataset.

- `tidy_map()` organizes the raw output of `map_ukbfgsea()` to a tibble including distinct columns for each phenotype and each ranking method, providing a clear and accessible summary of the FGSEA results.

## Installation

You can install the development version of ukbFGSEA like so:

``` r
# install.packages("devtools")
library(devtools)
install_github("ctlab/fgsea")
install_github("AzuleneG/ukbFGSEA")
```

`genebass_test_data` is a test data originated from `results.mt` provided on the Genebass website. The package implements it as a test data set for testing the functions. We includes the documentation of this dataset, but because this test data is very large (220.6MB), the file is only provided in the releases of repo `AzuleneG/ukbFGSEA`. You can install the data like so:

DO NOT install `piggyback` from CRAN to avoid a possible bug. Use `install_github()` instead.

```r
install_github("ropensci/piggyback")
library(piggyback)

piggyback::pb_download("genebass_test_data.rda",
	repo = "AzuleneG/ukbFGSEA",
	dest = tempdir())
load(paste0(tempdir(), "/genebass_test_data.rda"))
```

## Example

This example shows how to use `map_ukbfgsea()` to apply FGSEA on Genebass data, using the test data attached in the package.

``` r
library(tidyverse)
library(ukbFGSEA)

genebass_test_data 

genebass_test_data %>% 
  count(trait_type, phenocode, description, description_more)

# then apply ukbfgsea to three phenotypes for each of gene set
set.seed(123)
asd_output <- map_ukbfgsea(input_geneset = ASD185, multiple_phenotypes = genebass_test_data)

asd_output_tidy <- tidy_map(asd_output)
asd_output_tidy
```

## Tutorial


### Introduction

Welcome to the `ukbFGSEA` tutorial. This package is designed for performing fast pre-ranked gene set enrichment analysis (FGSEA) specifically tailored for the UK Biobank Genebass dataset. This tutorial aims to guide you through the effective use of the package, employing practical examples using gene sets related to Autism Spectrum Disorder (ASD), Developmental Delay (DD), and Neurodevelopmental Disorders (NDD).

### Practical Examples

#### Example 1: Quick Test Run

Since the Genebass data, including 4,529 phenotypes, is very large, running ukbFGSEA on all of them can take quite many hours. To illustrate how to run ukbFGSEA, we will use three phenotypes from Genebass as an example for a quick test run. Later, we will describe how to read in data, including all phenotypes from Genebass and apply ukbFGSEA to them.

```{r}
# First, load the libraries, Genebass Test Data, ASD Gene Set, DD Gene Set and NDD Gene Set
library(tidyverse)
library(ukbFGSEA)
library(piggyback) #Only for test data

# Download and load test data

pb_download("genebass_test_data.rda",
	repo = "AzuleneG/ukbFGSEA",
	dest = tempdir())
load(paste0(tempdir(), "/genebass_test_data.rda"))

# Display the structure of the test data
genebass_test_data %>%
  count(annotation,
        trait_type,
        phenocode,
        description,
        description_more)

# Apply ukbfgsea to three phenotypes for each gene set
set.seed(123)
asd_test_output <- map_ukbfgsea(input_geneset = ASD185, multiple_phenotypes = genebass_test_data)
asd_test_output_tidy <- tidy_map(asd_test_output)
asd_test_output_tidy

set.seed(123)
dd_test_output <- map_ukbfgsea(input_geneset = DD477, multiple_phenotypes = genebass_test_data)
dd_test_output_tidy <- tidy_map(dd_test_output)
dd_test_output_tidy

set.seed(123)
ndd_test_output <- map_ukbfgsea(input_geneset = NDD664, multiple_phenotypes = genebass_test_data)
ndd_test_output_tidy <- tidy_map(ndd_test_output)
ndd_test_output_tidy
```

#### Example 2: Applying ukbFGSEA on Genebass Data Including All Phenotypes

If you’d like to run ukbFGSEA on all phenotypes in the UK Biobank, first download the entire Genebass dataset from [Google Drive](https://drive.google.com/file/d/1j0bAXm2AYmZNQxQBbFODien0dpKUGCHn/view?usp=drive_link). Reading in the data may take several hours due to its large size. The runtime of ukbFGSEA depends on the number of phenotypes provided to the map_ukbfgsea function. Therefore, we suggest filtering out phenotypes that are not relevant to your interests and those with small sample sizes in the UK Biobank, as they lack power in burden tests. Generally, we recommend filtering out phenotypes with case sample sizes < 3000 and those with descriptions starting with “Date”. You are welcome to apply your own filtering criteria to clean the phenotypes.

In the following code example, we apply our recommended filters and focus on the burden test results of pLoF (protein-truncating) variants. Since it may take many hours to run ukbFGSEA on thousands of phenotypes, we provide the tidy output of map_ukbfgsea on filtered phenotypes in our package for users as an example. Users can simply load the data `asd_all_output_tidy`, `dd_all_output_tidy`, and `ndd_all_output_tidy`.
 

```{r eval=F, echo=T}
library(tidyverse)
library(ukbFGSEA)

all_phenotypes <- readRDS(file = "/the/path/to/genebass_all_phenotypes.rds")

phenotypes_filtered <- all_phenotypes %>%
  filter(n_cases >= 3000) %>%
  filter(!str_detect(description, "^Date ")) %>%
  filter(annotation == "pLoF")

set.seed(123)
asd_all_output <- map_ukbfgsea(input_geneset = ASD185, multiple_phenotypes = phenotypes_filtered)
asd_all_output_tidy <- tidy_map(asd_all_output)
asd_all_output_tidy

set.seed(123)
dd_all_output <- map_ukbfgsea(input_geneset = DD477, multiple_phenotypes = phenotypes_filtered)
dd_all_output_tidy <- tidy_map(dd_all_output)
dd_all_output_tidy

set.seed(123)
ndd_all_output <- map_ukbfgsea(input_geneset = NDD664, multiple_phenotypes = phenotypes_filtered)
ndd_all_output_tidy <- tidy_map(ndd_all_output)
ndd_all_output_tidy
```

### Results Interpretation

#### Identifying Phenotypes Significantly Enriched with Input Gene Set

To identify which phenotypes are significantly enriched with ASD, DD and NDD genes respectively, we apply a p-value cutoff and Enrichment Score (ES) filter. Here, we first load the results on filtered phenotypes with the burden test of pLoF variants as examples and apply the criteria.

```{r fig.width=9.11, fig.height=6.22}
# Load the tidy results
data("asd_all_output_tidy")
data("ndd_all_output_tidy")
data("dd_all_output_tidy")

#Function to extract the significant phenotypes 
extract_significant_phenotypes <- function(data) {
  data <- data %>%
    ungroup() %>%
    mutate(description_full = paste0(description, "; ", coding_description))
  
  cor_value <- cor(data$beta_pval, data$sign_pval, use = "complete.obs")
  print(paste("Correlation between beta_pval and sign_pval:", cor_value))
  
  cutoff <- 0.05 / nrow(data)
  
  data_sig <- data %>%
    filter(beta_pval <= cutoff, sign_pval <= cutoff,
           beta_NES > 0, sign_NES > 0) %>%
    mutate(description_full = fct_reorder(description_full, beta_pval))
  
  return(data_sig)
}

# extract phenotypes significantly enriched with ASD genes
significant_with_asd <- extract_significant_phenotypes(asd_all_output_tidy)

# extract phenotypes significantly enriched with DD genes
significant_with_dd <- extract_significant_phenotypes(dd_all_output_tidy)

# extract phenotypes significantly enriched with NDD genes
significant_with_ndd <- extract_significant_phenotypes(ndd_all_output_tidy)

significant_with_asd
significant_with_dd
significant_with_ndd


# An example to visualize phenotypes significantly enriched with ASD
significant_with_asd %>% 
  ggplot(aes(description_full, -log10(beta_pval)))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  xlab("")+
  ylab("-log10(P value from Beta metric)")

```

#### Biological Relevance

We discuss the phenotypes significantly enriched with ASD genes as example here. The results highlight the complex interplay between ASD genes and a wide array of phenotypic traits in the UK Biobank. These findings can be interpreted in several ways:

1. **Comorbidities and Genetic Overlap:** The significant enrichment of ASD genes in conditions like epilepsy, panic attacks, and various physical ailments underscores the multifaceted nature of ASD. It suggests that the genetic basis of ASD may also predispose individuals to other neurological and physical health conditions.

2. **Implications for Treatment and Management:** Understanding these genetic linkages can inform more comprehensive healthcare strategies for individuals with ASD. For instance, the link between ASD genes and metabolic conditions like diabetes (insulin product usage) can lead to integrated care approaches that address both neurodevelopmental and metabolic health.

3. **Behavioral and Lifestyle Factors:** The significant enrichment in alcohol consumption categories provides insights into the behavioral patterns associated with ASD. These patterns may be influenced by the genetic makeup of individuals and can have implications for lifestyle interventions and support mechanisms.
	
4. **Medical Interventions and Surgical Needs:** The variety of surgeries and medical procedures associated with enriched ASD genes indicate that individuals with ASD might have higher healthcare needs, particularly related to musculoskeletal and chronic conditions. This highlights the importance of specialized medical care and preventive strategies in this population.
	
5. **Potential for Further Research:** These findings open avenues for further research into the specific genetic mechanisms underlying these associations. Investigating the pathways and genes involved could lead to better diagnostic tools and targeted therapies for both ASD and its comorbid conditions.

In conclusion, the significant enrichment of ASD genes across diverse phenotypes in the UK Biobank underscores the broad impact of these genetic factors on health and disease. This analysis not only enhances our understanding of ASD but also informs holistic approaches to healthcare and research for individuals with ASD-related genetic profiles. Future studies should aim to unravel the specific genetic and molecular mechanisms driving these associations to develop more effective interventions and support systems.

We hope this tutorial helps you effectively use the `ukbFGSEA` package for your gene set enrichment analysis.

## Getting help

If you find problems, please report them with clear information on [GitHub](https://github.com/tidyverse/ukbFGSEA/issues). You can also contact the authors through email: `hedyzhu615@gmail.com` and `pengjun.guo@outlook.com`